---
title: Construct Patient Database
subtitle: Ayush Noori
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE}
# knitr::opts_knit$set(root.dir = normalizePath(""))
```

# Dependencies

Load requisite packages and define directories. Note that this script uses my personal utilities package `brainstorm`, which can be downloaded via `devtools::install_github("ayushnoori/brainstorm")`.

```{r load-packages, message=FALSE, warning=FALSE}

# data manipulation
library(data.table)
library(purrr)
library(magrittr)

# relative file paths
library(here)

# string manipulation and UUID generation
library(stringr)
library(uuid)

# R interface to Postgres
library(DBI)
library(RPostgres)

# utility functions
library(brainstorm)

```

Note that directories are relative to the R project path.

```{r define-directores}

# set directories
data_dir = here("patient_data")

```

Define reader function.

```{r file-reader}

# convenience function to print column names
print_colnames = function(data_table) {
  cat(paste0('"', paste0(colnames(data_table), collapse = '", "'), '"'))
}

# function to read and clean files
file_reader = function(file_name, label, camelCase = T) {
  
  # read table
  data_table = fread(here(file_name))
  
  # filter empty values and column names
  # data_table[data_table == ""] <- NA
  colnames(data_table) = str_to_lower(colnames(data_table))
  
  # if not camelCase, return already
  if (!camelCase) { return(data_table) }
  
  # apply conditional filters generated with
  # print_colnames(patient_data$TABLE)
  
  # add primary keys
  if(label %in% c("Allergies", "Conditions", "Devices", "Immunizations", "Medications", "Observations", "PayerTransitions", "Procedures", "Supplies")) {
    data_table[, id := UUIDgenerate(n = nrow(data_table), output = "string")]
    # data_table[, id := as.character(id)]
    setcolorder(data_table, "id")
  }
  
  # if ("patient" %in% colnames(data_table)) { data_table[, patient := as.character(patient)] }
  # if ("encounter" %in% colnames(data_table)) { data_table[, encounter := as.character(encounter)] }
  # if ("id" %in% colnames(data_table)) { data_table[, id := as.character(id)] }
  
  # no need to modify allergies, conditions, devices, observations, organizations, providers
  
  # filter
  if(label == "CarePlans") {
    setnames(data_table, c("reasoncode", "reasondescription"), c("reasonCode", "reasonDescription"))
    
  } else if(label == "Claims") {
    # note that I rename patientId and providerId to patient and provider
    setnames(data_table, colnames(data_table),
             c("id", "patient", "provider", "primaryPatientInsuranceId", "secondaryPatientInsuranceId", "departmentId", "patientDepartmentId", "diagnosis1", "diagnosis2", "diagnosis3", "diagnosis4", "diagnosis5", "diagnosis6", "diagnosis7", "diagnosis8", "referringProviderId", "appointmentId", "currentIllnessDate", "serviceDate", "supervisingProviderId", "status1", "status2", "statusp", "outstanding1", "outstanding2", "outstandingP", "lastBilledDate1", "lastBilledDate2", "lastBilledDateP", "healthcareClaimTypeId1", "healthcareClaimTypeId2"))
    
    # replace NA values
    data_table[primaryPatientInsuranceId == "0", primaryPatientInsuranceId := NA]
    data_table[secondaryPatientInsuranceId == "0", secondaryPatientInsuranceId := NA]
    data_table[, referringProviderId := as.character(referringProviderId)]
    
  } else if(label == "ClaimsTransactions") {
    # note that I rename patientId and providerId to patient and provider
    setnames(data_table, colnames(data_table),
             c("id", "claimId", "chargeId", "patient", "type", "amount", "method", "fromDate", "toDate", "placeOfService", "procedureCode", "modifier1", "modifier2", "diagnosisRef1", "diagnosisRef2", "diagnosisRef3", "diagnosisRef4", "units", "departmentId", "notes", "unitAmount", "transferOutId", "transferType", "payments", "adjustments", "transfers", "outstanding", "appointmentId", "lineNote", "patientInsuranceId", "feeScheduleId", "provider", "supervisingProviderId"))
    
  } else if(label == "Encounters") {
    setnames(data_table, c("encounterclass", "base_encounter_cost", "total_claim_cost", "payer_coverage", "reasoncode", "reasondescription"), c("encounterClass", "baseEncounterCost", "totalClaimCost", "payerCoverage", "reasonCode", "reasonDescription"))
    
  } else if(label == "ImagingStudies") {
    setnames(data_table, colnames(data_table),
             c("id", "date", "patient", "encounter", "seriesUid", "bodySiteCode", "bodySiteDescription", "modalityCode", "modalityDescription", "instanceUid", "sopCode", "sopDescription", "procedureCode"))
    
  } else if(label == "Immunizations") {
    setnames(data_table, "base_cost", "baseCost")
  
  } else if(label == "Medications") {
    setnames(data_table, colnames(data_table),
             c("id", "start", "stop", "patient", "payer", "encounter", "code", "description", "baseCost", "payerCoverage", "dispenses", "totalCost", "reasonCode", "reasonDescription"))
    
  } else if(label == "Observations") {
    data_table[encounter == "", encounter := NA]
    
  } else if(label == "Patients") {
    setnames(data_table, c("birthdate", "deathdate", "birthplace", "healthcare_expenses", "healthcare_coverage"), c("birthDate", "deathDate", "birthPlace", "healthcareExpenses", "healthcareCoverage"))
    
    # remove numbers from names
    data_table[, first := gsub("[0-9.]", "", first)]
    data_table[, last := gsub("[0-9.]", "", last)]
    data_table[, maiden := gsub("[0-9.]", "", maiden)]
    
  } else if(label == "PayerTransitions") {
    setnames(data_table, colnames(data_table), c("id", "patient", "memberId", "startDate", "endDate", "payer", "secondaryPayer", "planOwnership", "ownerName"))
    
    # replace NA values
    data_table[secondaryPayer == "", secondaryPayer := NA]
  
  } else if(label == "Payers") {
    setnames(data_table, colnames(data_table), c("id", "name", "ownership", "address", "city", "stateHeadquartered", "zip", "phone", "amountCovered", "amountUncovered", "revenue", "coveredEncounters", "uncoveredEncounters", "coveredMedications", "uncoveredMedications", "coveredProcedures", "uncoveredProcedures", "coveredImmunizations", "uncoveredImmunizations", "uniqueCustomers", "qolsAvg", "memberMonths"))
    
  } else if(label == "Procedures") {
    setnames(data_table, c("base_cost", "reasoncode", "reasondescription"), c("baseCost", "reasonCode", "reasonDescription"))
    
  }
  
  # return table
  return(data_table)
  
}

```

Read files and convert to camel case, which is the PostgreSQL standard.

```{r read-files}

# generate file list
patient_files = list.files(data_dir, include.dirs = F, full.names = T, pattern = ".csv")
new_names = gsub(".csv", "", basename(patient_files), fixed = T) %>%
  str_to_title() %>%
  strsplit("_") %>%
  map_chr(function(x) {ifelse(is.na(x[2]), x[1], paste0(x[1], str_to_title(x[2])))})
new_names[new_names == "Careplans"] <- "CarePlans"
names(patient_files) = new_names

# read files and convert to camel case
patient_data = imap(patient_files, ~file_reader(.x, .y, camelCase = T))
# count number of columns via map(patient_data, ~length(colnames(.x)))

```

The tables are specified as follows.

| File | Description |
|------|-------------|
| `allergies.csv` | Patient allergy data. |
| `careplans.csv` | Patient care plan data, including goals. |
| `claims.csv` | Patient claim data. |
| `claims_transactions.csv` | Transactions per line item per claim. |
| `conditions.csv` | Patient conditions or diagnoses. |
| `devices.csv` | Patient-affixed permanent and semi-permanent devices. |
| `encounters.csv` | Patient encounter data. |
| `imaging_studies.csv` | Patient imaging metadata. |
| `immunizations.csv` | Patient immunization data. |
| `medications.csv` | Patient medication data. |
| `observations.csv` | Patient observations including vital signs and lab reports. |
| `organizations.csv` | Provider organizations including hospitals. |
| `patients.csv` | Patient demographic data. |
| `payer_transitions.csv` | Payer Transition data (i.e. changes in health insurance). |
| `payers.csv` | Payer organization data. |
| `procedures.csv` | Patient procedure data including surgeries. |
| `providers.csv` | Clinicians that provide patient care. |
| `supplies.csv` | Supplies used in the provision of care. |

Write tables to PostgreSQL database.

```{r write-database}

# open connection
con = dbConnect(drv = RPostgres::Postgres(), dbname = "amanuensis")

# write tables
for (i in 1:length(patient_data)) {
  
  # subset table
  patient_data_table = patient_data[[i]]
  table_name = names(patient_data)[i]
  
  # write to database
  dbWriteTable(con, name = table_name, value = patient_data_table)
   
  # remove table
  # dbRemoveTable(con, name = table_name)
  
}

```

Execute SQL queries to create keys.

```{r sql-queries}

# write SQL queries to set keys
queries = list(
  
  # make primary keys first
  'ALTER TABLE "CarePlans" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Claims" ADD PRIMARY KEY (id);',
  'ALTER TABLE "ClaimsTransactions" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Encounters" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Organizations" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Patients" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Payers" ADD PRIMARY KEY (id);',
  # 'ALTER TABLE "PayerTransitions" ADD PRIMARY KEY ("memberId");', # added primary key
  'ALTER TABLE "Providers" ADD PRIMARY KEY (id);',
  
  # new primary keys: allergies, conditions, devices, immunizations, medications, observations, payerTransitions, procedures, supplies, imagingStudies
  'ALTER TABLE "Allergies" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Conditions" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Devices" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Immunizations" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Medications" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Observations" ADD PRIMARY KEY (id);',
  'ALTER TABLE "PayerTransitions" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Procedures" ADD PRIMARY KEY (id);',
  'ALTER TABLE "Supplies" ADD PRIMARY KEY (id);',
  'ALTER TABLE "ImagingStudies" ADD PRIMARY KEY (id);',
  
  # allergies
  'ALTER TABLE "Allergies" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  
  # carePlans
  'ALTER TABLE "CarePlans" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "CarePlans" ADD FOREIGN KEY (encounter) REFERENCES "Encounters"(id);',
  
  # claims
  'ALTER TABLE "Claims" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "Claims" ADD FOREIGN KEY (provider) REFERENCES "Providers"(id);',
  'ALTER TABLE "Claims" ADD FOREIGN KEY ("primaryPatientInsuranceId") REFERENCES "Payers"(id);',
  'ALTER TABLE "Claims" ADD FOREIGN KEY ("secondaryPatientInsuranceId") REFERENCES "Payers"(id);',
  'ALTER TABLE "Claims" ADD FOREIGN KEY ("referringProviderId") REFERENCES "Providers"(id);',
  'ALTER TABLE "Claims" ADD FOREIGN KEY ("appointmentId") REFERENCES "Encounters"(id);',
  'ALTER TABLE "Claims" ADD FOREIGN KEY ("supervisingProviderId") REFERENCES "Providers"(id);',
  
  # claimsTransactions
  'ALTER TABLE "ClaimsTransactions" ADD FOREIGN KEY ("claimId") REFERENCES "Claims"(id);',
  'ALTER TABLE "ClaimsTransactions" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "ClaimsTransactions" ADD FOREIGN KEY ("placeOfService") REFERENCES "Organizations"(id);',
  'ALTER TABLE "ClaimsTransactions" ADD FOREIGN KEY ("appointmentId") REFERENCES "Encounters"(id);',
  # 'ALTER TABLE "ClaimsTransactions" ADD FOREIGN KEY ("patientInsuranceId") REFERENCES "PayerTransitions"("memberId");',
  'ALTER TABLE "ClaimsTransactions" ADD FOREIGN KEY (provider) REFERENCES "Providers"(id);',
  'ALTER TABLE "ClaimsTransactions" ADD FOREIGN KEY ("supervisingProviderId") REFERENCES "Providers"(id);',
  
  # conditions
  'ALTER TABLE "Conditions" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "Conditions" ADD FOREIGN KEY (encounter) REFERENCES "Encounters"(id);',
  
  # devices
  'ALTER TABLE "Devices" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "Devices" ADD FOREIGN KEY (encounter) REFERENCES "Encounters"(id);',
  
  # encounters
  'ALTER TABLE "Encounters" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "Encounters" ADD FOREIGN KEY (organization) REFERENCES "Organizations"(id);',
  'ALTER TABLE "Encounters" ADD FOREIGN KEY (provider) REFERENCES "Providers"(id);',
  'ALTER TABLE "Encounters" ADD FOREIGN KEY (payer) REFERENCES "Payers"(id);',
  
  # imagingStudies
  'ALTER TABLE "ImagingStudies" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "ImagingStudies" ADD FOREIGN KEY (encounter) REFERENCES "Encounters"(id);',
  
  # immunizations
  'ALTER TABLE "Immunizations" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "Immunizations" ADD FOREIGN KEY (encounter) REFERENCES "Encounters"(id);',
  
  # medications
  'ALTER TABLE "Medications" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "Medications" ADD FOREIGN KEY (payer) REFERENCES "Payers"(id);',
  'ALTER TABLE "Medications" ADD FOREIGN KEY (encounter) REFERENCES "Encounters"(id);',
  
  # observations
  'ALTER TABLE "Observations" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "Observations" ADD FOREIGN KEY (encounter) REFERENCES "Encounters"(id);',
  
  # payerTransitions
  'ALTER TABLE "PayerTransitions" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "PayerTransitions" ADD FOREIGN KEY (payer) REFERENCES "Payers"(id);',
  'ALTER TABLE "PayerTransitions" ADD FOREIGN KEY ("secondaryPayer") REFERENCES "Payers"(id);',
  
  # procedures
  'ALTER TABLE "Procedures" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "Procedures" ADD FOREIGN KEY (encounter) REFERENCES "Encounters"(id);',
  
  # providers
  'ALTER TABLE "Providers" ADD FOREIGN KEY (organization) REFERENCES "Organizations"(id);',
  
  # supplies
  'ALTER TABLE "Supplies" ADD FOREIGN KEY (patient) REFERENCES "Patients"(id);',
  'ALTER TABLE "Supplies" ADD FOREIGN KEY (encounter) REFERENCES "Encounters"(id);'
  
)

```

Execute queries.

```{r execute-queries}

# walk across and execute queries
for (query in queries){
  data_query = dbSendQuery(con, query)
  dbClearResult(data_query)
}
  
# disconnect from the database
dbDisconnect(con)

```